---
title: "FPM #3: Drafting Visualizations"
subtitle: "Assigned Wed 02/11/2026 | Due Sun 02/22/2026"
format: html
author: "Austin Martinez"
date: 02/22/2026
---

```{r}
# Load in libraries
library(here)
library(tidyverse)
library(ggplot2)
library(ggimage)
library(sf)
library(tmap)
library(tigris)
library(dplyr)
library(stringr)
```

```{r}
# Read in Data
lake_data <- read_csv(here("data/EPA_EMAP_CHEM.csv"))
```

    -   What lakes in the Northeast United States have a mean nitrate (NO3) concentration exceeding the EPA water quality standard of 10 mg/L?

```{r fig.height = 7, fig.width=15}

# Find the mean NH4, NO3, NTL, PTL, CHLA, CL, PHEQ, TURB value for each lake
grouped_2 <- lake_data |>
  group_by(LAKENAME) |>
  summarize(
    mean_NH4  = mean(NH4,  na.rm = TRUE),
    mean_NO3  = mean(NO3,  na.rm = TRUE),
    mean_NTL  = mean(NTL,  na.rm = TRUE),
    mean_PTL  = mean(PTL,  na.rm = TRUE),
    mean_CHLA = mean(CHLA, na.rm = TRUE),
    mean_CL   = mean(CL,   na.rm = TRUE),
    mean_PHEQ = mean(PHEQ, na.rm = TRUE),
    mean_TURB = mean(TURB, na.rm = TRUE),
    LAT_DD = mean(LAT_DD, na.rm = TRUE),
    LON_DD = mean(LON_DD, na.rm = TRUE)
  )

# Convert units to match EPA WQS

# NH4: microequivalentsPerLiter to mg/L
grouped_2$mean_NH4 <- grouped_2$mean_NH4 * 0.01804
# NO3: microequivalentsPerLiter to mg/L
grouped_2$mean_NO3 <- grouped_2$mean_NO3 * 0.062
# NTL: microgramsPerLiter to mg/L
grouped_2$mean_NTL <- grouped_2$mean_NTL / 1000
# PTL: microgramsPerLiter to mg/L
grouped_2$mean_PTL <- grouped_2$mean_PTL / 1000
# CL: microequivalentsPerLiter to mg/L
grouped_2$mean_CL <- grouped_2$mean_CL * 0.03545


# Add EPA WQS threshold columns
grouped_2$wqs_NH4   <- 0.1      # mg/L
grouped_2$wqs_NO3   <- 10       # mg/L
grouped_2$wqs_NTL   <- 1        # mg/L
grouped_2$wqs_PTL   <- 0.10     # mg/L
grouped_2$wqs_CHLA  <- 5        # μg/L 
grouped_2$wqs_CL    <- 230      # mg/L
grouped_2$wqs_PHEQ  <- 7.0      # pH 
grouped_2$wqs_TURB  <- 5        # NTU

# Count violations for each lake
grouped_2$violations <- 
  (grouped_2$mean_NH4 >= grouped_2$wqs_NH4) +
  (grouped_2$mean_NO3 >= grouped_2$wqs_NO3) +
  (grouped_2$mean_NTL >= grouped_2$wqs_NTL) +
  (grouped_2$mean_PTL >= grouped_2$wqs_PTL) +
  (grouped_2$mean_CHLA >= grouped_2$wqs_CHLA) +
  (grouped_2$mean_CL >= grouped_2$wqs_CL) +
  ((grouped_2$mean_PHEQ < 6.5) | (grouped_2$mean_PHEQ > 8.5)) +
  (grouped_2$mean_TURB >= grouped_2$wqs_TURB)

# NH4: microequivalentsPerLiter to mg/L
grouped_2$mean_NH4 <- grouped_2$mean_NH4 * 0.01804

# Find the mean NO3 value for each lake
grouped <- lake_data |>
  group_by(LAKENAME) |>
  summarize(mean_NO3 = mean(NO3))

# Convert units to mg/L
grouped$mean_NO3 <- grouped$mean_NO3 * 0.062
```



```{r fig.height = 24, fig.width=11}
# Find mean NO3 value for each lake
grouped <- lake_data |>
  group_by(LAKENAME) |>
  summarize(mean_NO3 = mean(NO3, na.rm = TRUE)) |>
  mutate(
    mean_NO3 = mean_NO3 * 0.062,
    LAKENAME = str_to_sentence(LAKENAME)
  ) |>
  arrange(desc(mean_NO3)) |>
  slice_head(n = 6)

# Make lake order numeric for wave plotting
grouped$LAKENAME <- factor(grouped$LAKENAME, levels = rev(grouped$LAKENAME))
grouped$x <- as.numeric(grouped$LAKENAME)

# Make bars wavy
wave_data <- grouped |>
  rowwise() |>
  do({
    y_vals <- seq(0, .$mean_NO3, length.out = 200)
    data.frame(
      LAKENAME = .$LAKENAME,
      y = y_vals,
      x = .$x + 0.1 * sin(seq(0, 5*pi, length.out = 200))
    )
  })

# Plot
NO3 <- ggplot(grouped, aes(x = x, y = mean_NO3)) +
  geom_path(
    data = wave_data,
    aes(x = x, y = y, group = LAKENAME),
    color = "darkgreen",
    linewidth = 5
  ) +

  # value labels
  geom_text(
    aes(label = signif(mean_NO3,3)),
    vjust = -.8,
    size = 9
  ) +

  # EPA WQS line
  geom_hline(
    yintercept = 10,
    linetype = "dashed",
    linewidth = 1
  ) +

  scale_x_continuous(
    breaks = grouped$x,
    labels = grouped$LAKENAME
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +

  labs(
    x = "Lakes",
    y = "Mean NO3 concentration"
    #title = "Lakes with Mean NO3 Concentrations Exceeding EPA Water Quality Standards"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.title.y = element_text(size = 28, face = "bold"),
    axis.title.x = element_blank(),
    axis.text.y = element_text(size = 30, face = "bold", color = "black"),
    axis.text.x = element_text(size = 24, face = "bold", color = "black", angle = 37, hjust = 1),
    
    panel.grid.major.x = element_blank(),   # remove x axis gridlines
    panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank(),

    text = element_text(size = 18),
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 22, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )

  # Save the plot
ggsave(
  plot = NO3,
  filename = "NO3_bar.png",
  width = 11,
  height = 24.3,
  units = "in",
  dpi = 300,
  bg = "transparent"
)

# View the plot
NO3

```



-   Which lakes have the greatest number of contaminants exceeding EPA Water Quality Standards?


```{r fig.height = 3, fig.width = 12}
# Read dead fish image
skull_url <- here::here("images/skelleton.png")

EPA_line <- grouped_2 |>
  arrange(desc(violations)) |>
  slice_head(n = 6) |>
  mutate(
    LAKENAME = str_to_sentence(gsub("\\s*\\(largest\\)", "", LAKENAME, ignore.case = TRUE)),
    skull_img = skull_url
  ) |>
  ggplot(aes(x = reorder(LAKENAME, violations),
             y = violations)) +
  geom_linerange(aes(ymin = 1, ymax = violations),
                 color = "#41CBE2",
                 linewidth = 2) +
  geom_image(aes(image = skull_img), size = 0.4) +
  geom_text(aes(label = violations, hjust = -4.5), size = 5.5) +
  labs(
    x = "Lakes",
    y = "Number of EPA WQS Violations") +
    #title = "Lakes with Most EPA Water Quality Standards (WQS) Violations",
    #subtitle = "Contaminants examined: NH4, NO3, Cl, Total Nitrogen, Total Phosphorus, Chlorophyll A, pH, Turbidity") +
  coord_flip() +
scale_y_continuous(
  limits = c(1, max(grouped_2$violations, na.rm = TRUE)),
  breaks = 1:7,
  expand = expansion(mult = c(0, .31))
) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 18, face = "bold", color = "black"),
    axis.text.x = element_text(size = 18, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(),
    panel.grid.minor.x = element_blank(),
    text = element_text(size = 18),
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 22, face = "bold"),
    plot.subtitle = element_text(size = 13),
    plot.caption = element_text(size = 9),
    plot.background  = element_rect(fill = NA, color = NA),
    panel.background = element_rect(fill = NA, color = NA)
  )

  # Save the plot
ggsave(
  plot = EPA_line,
  filename = "EPA_line.png",
  width = 12,
  height = 3,
  units = "in",
  dpi = 300,
  bg = "transparent"
)

# View plot
EPA_line

```



    -   Where are the lakes that have contaminant concentrations above the EPA Water Quality Standards located? Are they close together and possibly being contaminated by the same source?



```{r}
# Gather aberviations
states_ne <- c("ME","NH","VT","MA","RI","CT","NY","NJ","PA")

states_sf <- states(year = 2023) |>
  filter(STUSPS %in% states_ne) |>
  st_as_sf()
```

```{r fig.height = 30, fig.width= 30}
# Add manual offsets for each state abbreviation
states_sf <- states_sf %>%
  mutate(
    x_offset = case_when(
      STUSPS == "ME" ~ 0.9,
      STUSPS == "NH" ~ 0.5,
      STUSPS == "VT" ~ 0,
      STUSPS == "MA" ~ 0,
      STUSPS == "RI" ~ 0.55,
      STUSPS == "CT" ~ 0,
      STUSPS == "NY" ~ -0.5,
      STUSPS == "NJ" ~ 0.4,
      STUSPS == "PA" ~ 0,
      TRUE ~ 0
    ),
    y_offset = case_when(
      STUSPS == "ME" ~ 0.5,
      STUSPS == "NH" ~ 0.17,
      STUSPS == "VT" ~ 0.3,
      STUSPS == "MA" ~ 0.2,
      STUSPS == "RI" ~ -0.2,
      STUSPS == "CT" ~ -0.0,
      STUSPS == "NY" ~ 0,
      STUSPS == "NJ" ~ -0.3,
      STUSPS == "PA" ~ -0.2,
      TRUE ~ 0
    )
  )

```

```{r fig.height = 30, fig.width= 30}

# Convert grouped_2 to spatial object
grouped_2_sf <- st_as_sf(grouped_2, 
                         coords = c("LON_DD", "LAT_DD"), 
                         crs = 4326)

# Rename violations column and make it an integer
grouped_2_sf <- grouped_2_sf %>%
  rename(`Number of Violations` = violations) %>%
  mutate(`Number of Violations` = factor(`Number of Violations`))

ne_outline <- states_sf |>
  st_union() |>
  st_as_sf()

NE_map <- tm_shape(states_sf) +
  tm_polygons(col = "#C7F5FC", border.col = "#41CBE2", lwd = 4) +

  tm_shape(ne_outline) +
  tm_borders(col = "black", lwd = 10) +   # Make outer regional outline bold black

  tm_shape(states_sf) +
  tm_text(
    "STUSPS",
    size = 3,
    col = "black",
    fontface = "bold",
    xmod = "x_offset",
    ymod = "y_offset"
  ) +

  tm_shape(grouped_2_sf) +
  tm_dots(
    col = "Number of Violations",
    size = 1.3,
    palette = c("green", "yellowgreen", "gold", "orange", "red", "firebrick4", "black")
  ) +

  tm_layout(
    frame = FALSE,
    legend.show = FALSE,
    bg.color = "transparent"
  ) 

tmap_save(
  tm = NE_map,
  filename = "NE_map.png",
  width = 30,   
  height = 30,
  units = "in",
  dpi = 300,     # resolution
  bg = "transparent" # Take away background
)

# Show map
NE_map

```


